---
title: "Uva Island bleaching"
author: "Ana M. Palacio-Castro"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
    theme: united
    code_folding: "hide"
bibliography: packages.bib
nocite: '@*'
---

# Setup

```{r Knitr_setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	fig.width=8, fig.height=8
)
```


```{r libraries, include=FALSE}

library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(nlme)
library(multcomp)
library(emmeans)
library(gridExtra)
library(lubridate)

theme_set (theme_classic() + theme(panel.grid.major = element_blank(),
                              panel.grid.minor = element_blank(), 
                              axis.line = element_line(colour = "black"),
                              legend.position="none",
                              axis.text.x = element_text(angle = 90, vjust = 0.5),
                              plot.title = element_text(size=12, face="bold"),
                              #panel.border = element_rect(colour = "black", fill=NA, size=1)
                              panel.border = element_blank()
                              ))
```

# Coral bleaching data 

## Import data 

### 1981-2016 data


```{r}
# Read data 
  Bleaching_proportion <- read.csv(
    "BleachingData/All_Proportion_Long2.csv", header=TRUE, sep = ",")
  summary(Bleaching_proportion)
  
  Bleaching_proportion<-Bleaching_proportion %>% drop_na(value)
  
  Bleaching_proportion$Date<-as.Date(Bleaching_proportion$Date,
                                     format="%m/%d/%y")
  Bleaching_proportion$Year<-year(Bleaching_proportion$Date)
  Bleaching_proportion$Date_F<-as.factor(Bleaching_proportion$Date)
  
  Bleaching_proportion$Genus<-factor(Bleaching_proportion$Genus,
                              levels = c("Millepora",
                                         "Other scleractinians", "Pocillopora"))
  
  #Bleaching_proportion$Date_F<-factor(Bleaching_proportion$Date,
  #                                    levels=c("20-03-1983", "22-03-1983",
  #                                            "27-04-1983", "15-03-1998",
  #                                            "15-08-2014", "15-08-2015",
  #                                            "15-04-2016"))
  Bleaching_proportion<-subset(Bleaching_proportion, Date_F!="2014-08-15")
  
  Bleaching_general<-subset(Bleaching_proportion, Category=="General")
  Bleaching_spp<-subset(Bleaching_proportion, Category=="Species")
  

    
```

```{r}
Plot_all_boxplot <- ggplot(Bleaching_proportion, aes(
                        x=as.factor(Date_F), y=value)) +
  geom_boxplot(outlier.shape = NA) +
  stat_boxplot(geom = 'errorbar')+
   geom_jitter(aes(fill=Species), shape=21, alpha=0.8, width = 0.15) +
    #stat_summary(fun.data = mean_se, geom = "errorbar")+
    #stat_summary(fun=mean, geom="point", size=2) +
    #stat_summary(fun="mean_cl_normal", geom = "errorbar", size=2) +
    #stat_summary(fun.y=median, geom="point", size=2, shape=14) +
  facet_grid(Genus~Condition) + theme(legend.position = "right") +
  xlab("Date") + ylab("Proportion of corals")
Plot_all_boxplot

#ggsave(file="Outputs/Bleaching2.svg", plot=Plot_all_boxplot2, width=6, height=5)
```


```{r}
Plot_all <- ggplot(Bleaching_proportion, aes(
                        x=as.factor(Year), y=value)) +
  geom_boxplot(outlier.shape = NA) +
  stat_boxplot(geom = 'errorbar')+
  geom_jitter(aes(fill=Species), shape=21, alpha=0.8, width = 0.15) +
  #geom_jitter(shape=21, width = 0.15) +
  stat_summary(fun=mean, geom="point", size=2) +
  theme(legend.position = "right") +
  xlab("Date") + ylab("Proportion of corals")+
   scale_color_hue(l=40, c=35)
Plot_all +  facet_grid(Condition~Genus)

#ggsave(file="Outputs/Bleaching_ENSO_pooled.svg", plot=Plot_all, width=7, height=6)
```


```{r}
Plot_ssp_g <- ggplot(Bleaching_spp, aes(
                        x=Date_F, y=value)) +
  geom_boxplot(outlier.shape = NA) +
  stat_boxplot(geom = 'errorbar')+
  geom_jitter(aes(fill=Species), shape=21, alpha=0.8, width = 0.15) +
  stat_summary(fun=mean, geom="point", size=2) +
  stat_summary(fun=mean, geom="line")+
  facet_grid(Genus~Condition) + theme(legend.position = "right") +
  xlab("Dates?") + ylab("Proportion of corals")
Plot_ssp_g 
#ggsave(file="Outputs/Bleaching_ENSOv2.svg", plot=Plot_ssp_g, width=7, height=6)
```


## Summary stats per event 

```{r, echo=FALSE}

# mean proportion of coral condition
Condition_Mean <- aggregate(value ~ 
                            Genus + Date_F+ Condition,
                            FUN=mean,
                          data=Bleaching_proportion)

# medain proportion of coral condition
Condition_Median <- aggregate(value ~ 
                           Genus + Date_F+ Condition,
                            FUN=median,
                          data=Bleaching_proportion)

Condition_SD <- aggregate(value ~ 
                           Genus + Date_F+ Condition,
                            FUN=sd,
                          data=Bleaching_proportion)

Condition_summary<-cbind(Condition_Mean, 
                         Condition_Median[,-1:-3], 
                         Condition_SD[,-1:-3])
  colnames(Condition_summary)<-c("Taxon", "Date","Condition", "Mean","Median", "SD")
  Condition_summary
  
  #write.csv(Condition_summary, "Outputs/Condition_summary.csv", row.names = F)
```
 
# Heat stress

```{r}
Location_date<-expand.grid(Location=unique(Bleaching_proportion$Location) ,
                        Date=unique(Bleaching_proportion$Date))
Dates<-expand.grid(Date=unique(Bleaching_proportion$Date))

```


# 1. SST Data 

### Extract temperature data based on the coordinates

```{r libraries1, message=FALSE, results="hide"}
  #library(raster)
  #library(parallel)
```
 
```{r Extract_SST_data, cache=TRUE}
   # read all the .nc files 
    #flist<-list.files('Daily_OISST_v2',pattern='*.nc',full.names=TRUE)
      
  # brick them 
    #ras<-lapply(flist, stack)
    pts <- read.csv("BleachingData/Coordinates.csv", stringsAsFactors = FALSE)
    #extract.pts <- cbind(lon=pts$lon, lat=pts$lat)
    
  ## Extract data by coordinates 
    #ext <- lapply(ras,raster::extract,extract.pts)
    #head(ext)
    #SST <- unlist(ext, recursive = FALSE)
    #head(SST)
```


### Bind temperature data with coordinates and dates

```{r Bind_SST_GPS_Date}
  
# Create a date list 
    Dates1<-seq.Date(as.Date("1982-01-01"), as.Date("1983-12-31"), by = "day")
    Date1<-as.character(Dates1)
    
    Dates2<-seq.Date(as.Date("1985-01-01"), as.Date("2012-12-31"), by = "day")
    Date2<-as.character(Dates2)
    
    Dates3<-seq.Date(as.Date("2014-01-01"), as.Date("2016-12-31"), by = "day")
    Date3<-as.character(Dates3)
        
    Date<-c(Date1, Date2, Date3)
    #head(Date)
      
# Bind GPS points and Dates
    Date.location <- merge(pts, Date)
    #Temperature.data<-data.frame(Date.location, SST)
    #Temperature.data$Date<-as.Date(Temperature.data$y)
    # head(Temperature.data)  
    # summary(Temperature.data)
    # write.csv(Temperature.data, "Outputs/OISST_Temperature.csv")
```

# 2. DHW calculation

```{r libraries2, results="hide"}
  library(lubridate)
  library(dplyr)
  library(tidyverse)
  library(zoo)
  library(reshape2)
```

### Calculate MMM

For the climatology and maximum monthly mean (MMM) we selected the SST data from 1985-2012, to keep consistency with the CRW methodology

```{r Climatology}
  # Temperature.data$Month<-month(Temperature.data$Date)
  # Temperature.data$Year<-year(Temperature.data$Date)
  # Temperature.data$Location<-as.factor(Temperature.data$Location) 
  # 
  # # SELECT YEARS 1985-2012
  #   SST.1985_2012<-filter(Temperature.data, Year>1984)
  #   SST.1985_2012<-filter(SST.1985_2012, Year<2013)
  # # Mean temperature per month in each location
  # OI.MonthlyClimatology_1985_2012<-SST.1985_2012 %>%
  #   group_by(Location, Month) %>%
  #   summarize(Clima_M = mean(SST, na.rm = TRUE))
  # #OI.MonthlyClimatology_1985_2012
  # # Maximum Monthly Mean in each location
  # MMM_OI<-OI.MonthlyClimatology_1985_2012 %>%
  #   group_by(Location) %>%
  #   summarize(MMM = max(Clima_M, na.rm = TRUE))
  # # MMM_OI
```

### Calculate HotSpots

```{r HotSpots}
# Temperature.data <- merge(Temperature.data, MMM_OI, on="Location") # Add the MMM to each location
# Temperature.data <- Temperature.data[order(Temperature.data$Location, Temperature.data$Date),]
# Temperature.data$HotSpot<-(Temperature.data$SST-(Temperature.data$MMM))
# Temperature.data$HotSpot<-ifelse(Temperature.data$HotSpot>=0,
#                                 Temperature.data$HotSpot, 0)# REMOVE NEGATIVE ANOMALIES
# Temperature.data$D_Stress<-ifelse(Temperature.data$HotSpot>=1,
#                                Temperature.data$HotSpot, 0) # REMOVE HotSpots <1
# Temperature.data$W_Stress<-(Temperature.data$D_Stress/7) # Transform daily to weekly values
```

### Calculate DHW

```{r DHW}
# Rolling sum of 84 days (12 weeks) Hotspots for each location

# DHW_Daily<-Temperature.data %>% 
#   group_by(Location) %>%
#   mutate(NOAA_DHW_OI=rollapplyr(W_Stress,list(-(83:0)),sum,fill=NA, partial=FALSE)) %>%
#   ungroup
#write.csv(DHW_Daily, "DHW_Daily.csv")
DHW_Daily<-read.csv("BleachingData/DHW_Daily.csv")

DHW_month<-DHW_Daily %>% 
   group_by(Location) %>%
   mutate(DHW_max=rollapplyr(NOAA_DHW_OI,list(-(30:0)),
                  max,fill=NA, partial=FALSE)) %>% ungroup

 DHW_month<-as.data.frame(DHW_month)
 
 DHW_month<- DHW_month %>% dplyr::select("Location", "Date","NOAA_DHW_OI", "DHW_max")
 Date.location<- Date.location %>% dplyr::select("Location", "y")
 colnames(Date.location)<-c("Location", "Date")

```

```{r}
Location_date<-merge(Location_date, DHW_month, 
                     by=c("Location", "Date"), all.x=TRUE)
Bleaching_proportion<-merge(Bleaching_proportion, Location_date,
                            by=c("Location", "Date"), all.x=TRUE)
```


### Summarise max DHW and max SST by location and year

```{r}
# 1. Summary by Site and year (max values)
  DHW_Year_Summary<-DHW_Daily %>%
         group_by(Location, Year) %>%
         dplyr::summarise(maxDHW = max(NOAA_DHW_OI),
                          maxSST = max(SST))
  #write.csv(DHW_Year_Summary, "maxDHW.csv")

```

```{r}
DHW_loc <- ggplot(DHW_Daily, aes(
                        x=Date, y=NOAA_DHW_OI, colour=Location)) +
  geom_line() +
  xlab("Date") + ylab("DHW")+
  facet_grid(Location~.)+
  geom_vline(xintercept = c(Dates$Date))
  
DHW_loc 
```


```{r}
summary(Bleaching_proportion)

Summary_plot<-Bleaching_proportion %>%
         group_by(Year) %>%
         dplyr::summarise(meanDHW = mean(DHW_max),
                          seDHW = sd(DHW_max),
                          minDHW = min(DHW_max),
                          maxDHW = max(DHW_max))

Summary_plot2<-Bleaching_proportion %>%
         group_by(Year, Condition, Genus) %>%
         dplyr::summarise(meanCon = mean(value),
                          sdCon = sd(value),
                          minCon = min(value),
                          maxCon = max(value))
Sumeryplot<-merge(x = Summary_plot2, y = Summary_plot, by = "Year", all.x = TRUE)
Sumeryplot
```

```{r}
Plot<-ggplot(data=Sumeryplot,
      aes(x=meanDHW, y=meanCon, colour=factor(Year))) +
  #scale_colour_manual(values = c("magenta","cyan", "black", "black"), name="ENSO")+
  theme_bw()+
  theme(legend.position = "top",
        panel.grid= element_blank())+
       geom_point()+facet_grid(Condition~Genus)+
  #geom_jitter(data=Bleaching_proportion, aes(x=DHW_max, y=value))+
  scale_x_continuous(name="Max DHW",
                     limits = c(2,8),
                     breaks = c(2, 4, 6, 8))+
  geom_errorbarh(aes(
    xmax = meanDHW + seDHW, xmin = meanDHW - seDHW, colour=factor(Year)))+
  geom_errorbar(aes(
    ymax = meanCon + sdCon, ymin = meanCon - sdCon, colour=factor(Year)))
Plot
#ggsave(file="Outputs/Figure2_2.svg", plot=Plot, width=5, height=4)
```

```{r}
DHW_ENSO <- ggplot(data=Bleaching_proportion, aes(
                        x=DHW_max, y=value, shape=Genus)) +
  theme_bw()+
  theme(legend.position = "none",
        panel.grid= element_blank())+
  geom_jitter(aes(fill=Event, shape=Genus)) +
  scale_fill_manual(values = c("magenta","cyan", "black"), name="ENSO")+
  scale_shape_manual(values = c(21, 24, 22), name="ENSO")+
  facet_grid(Condition~Genus)

  
Year_ENSO <- ggplot(Bleaching_proportion, aes(
                        x=as.factor(Year), y=value, shape=Genus)) +
  #scale_x_continuous(limits = c(1982,2017), expand = c(0, 0), 
  #                   name=("Year"),
  #                   breaks = c(1983, 1998, 2015, 2016))+
   geom_jitter(aes(colour=Species), width = 0.15) +
  theme_bw()+
  theme(legend.position = "right",
        panel.grid= element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.text=element_text(size=8))+
  stat_summary(fun.data = "mean_cl_boot", colour = "gray")+
 
  #geom_jitter(aes(fill=Event, shape=Species)) +
  #geom_smooth(method = "lm", se=FALSE)+
  #scale_fill_manual(values = c("magenta","cyan", "black"), name="ENSO")+
  #scale_shape_manual(values = c(21, 22, 24), name="ENSO")+
  facet_grid(Condition~.)
Year_ENSO

Figure2<-grid.arrange(DHW_ENSO, Year_ENSO, ncol = 2)
 #ggsave(file="Outputs/Figure2.svg", plot=Figure2, width=9, height=4)


```

# Packages used

```{r}
# Creates bibliography 
# knitr::write_bib(c(.packages()), "packages.bib")
```
